Optimal Finite Element Integration

### 2015-08-03 TOMS_Paper/assembly.tex
Write optimization of Li quantities

In this section, we characterize our definition of optimality and we describe the assumptions under which it holds.

In order to make the document self-contained, we start by reviewing basic compiler terminology.

@@

Note that nothing prevents a perfect multilinear loop nest from being rooted in a deeper, possibly imperfect loop nest. In fact, we will later focus on this particular structure. First, we need to define optimality for perfect multilinear loop nests. To this purpose, it is convenient to introduce the notion of sharing.

@@

We can now present a simple yet fundamental result.

@@

Based on this proposition, we define optimality as follows.

@@

In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop. Note that this definition does not take into account memory requirements. If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense. In the following, we assume to operate in a CPU-bound regime, in which the body of loop nests are characterized by arithmetic-intensive expressions. This suits the context of finite element integration.

A second result follows.
\begin{Prop}
\label{prop:optimal-mln}
An optimal synthesis for a perfect multilinear loop nest $LN$ can deterministically be found.
\end{Prop}
\begin{proof}
By construction. Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$. These sets are sorted in descending order based on their cardinality. By establishing a one-to-one mapping between set indices and loop indices, we produce a new loop permutation. The loop permutation is semantically correct because of perfectness. In this new order, loops are placed such that as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$. Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}. This renders $L_{i_{n-1}}$ optimal. In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of the summation.
\end{proof}

We now consider generic loop nests in which multilinearity applies to a subset of loops only. Consider the example in Figure~\ref{code:loopnest}.

@@

\section{Synthesis of Optimal Loop Nests in Finite Element Integration}
\label{sec:optimal-syntesis}

@@

Eliminating sharing from the multilinear loops does not suffice. In fact, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, as already suggested in~\ref{def:optimality}, and the operation count be reduced.

@@

\begin{figure}[h]
\begin{lstlisting}[basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate $\mbox{\texttt{ops}}$ after pre-evaluation
  $\theta_{wo}$ = estimate $\mbox{\texttt{ops}}$ without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the main algorithm}
\label{code:intuition}
\end{figure}

@@

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. As described in Proposition~\ref{prop:optimal-mln}, a visit of the loop nest suffices to obtain the cost of the optimal synthesis of a multilinear loop nest, namely $MLN_{ops}$. Then, assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = MLN_{ops} \cdot I$.

@@

One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$. In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from other (reference element specific) quantities. By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient). $\iota$ serves to capture this growth.

@@

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$. In particular, we need to calculate the number of reference-element specific terms, $\rho$, such that $\theta_{w} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is characterized by just the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators employed. For example, in the case of the bi-dimensional Poisson equation, $\rho = 3$. There are several ways of determining $\rho$. The fastest would be to extract it from high-level analysis of the operators in the form; for convenience, in our implementation we instead project (by analysis of the expression tree) the output of expansion and factorization in the monomial.

@@

\section{Code Generation}
\label{sec:codegen}
The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake. In this section, we describe the structure of the code generation system and we comment on a set of low-level optimizations.

@@

\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
As a proof-of-concept of our generality claim,

@@

% MAYBE JUST A SUBSECTION, "INSIGHTS OF LOW-LEVEL OPTS ...."

>>>>

In this section, we provide a characterization of optimality for loop nests of various nature. In order to make the document self-contained, we start with reviewing basic compiler terminology.

@@

Note that perfect multilinear loop nests could actually be rooted in deeper, possibly imperfect loop nests. In fact, we will focus on this particular structure.

We introduce the fundamental notion of sharing.

@@

We now have the ingredients to formulate a simple yet fundamental result.

@@

This result will be essential to prove that optimality, characterized as follows, can structurally be reached in this class of loops.

@@

In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop. Note that this definition does not take into account memory requirements. If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense. In the following, we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated. This suits the context of finite element integration.

A second result follows.
\begin{Prop}
\label{prop:optimal-mln}
A perfect multilinear loop nest $LN$ can always be reduced to optimal form.
\end{Prop}
\begin{proof}
By construction. Loop-dependent terms are logically grouped into $n$ disjoint sets $S_i$, each $S_i$ containing all terms depending on $L_i$. These sets are sorted in descending order based on their cardinality. By establishing a one-to-one mapping between set indices and loop indices, we produce a new loop permutation. The loop permutation is semantically correct because of perfectness. In this new order, loops are placed such that as going down the nest, $L_i$ is characterized by less unique terms than $L_{i-1}$. Starting from $L_{i_{n-1}}$, we can apply the sharing-removal procedure described in Proposition~\ref{prop:share-removal}. This renders $L_{i_{n-1}}$ optimal. In particular, the number of operations is equal to $LN_{ops} = \# S_{i_{n-1}} + (\# S_{i_{n-1}} - 1)$, in which the second term represents the cost of summing the $\# S_{i_{n-1}}$ terms.
\end{proof}

For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_optimal}.

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}
\end{CenteredBox}\caption{Multilinear loop nest in optimal form}\label{code:multi_loopnest_optimal}\end{figure}

We now consider a wider class of loop nests in which perfectness and multilinearity apply to a sub-nest only. Consider the example in Figure~\ref{code:loopnest}.

@@

\section{Synthesis of Optimal Loop Nests in Finite Element Integration}
\label{sec:optimal-synthesis}

@@

Eliminating sharing from the inner multilinear loops does not suffice. In fact, as suggested in Section~\ref{def:optimality}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

@@

\begin{figure}[h]
\begin{lstlisting}[basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the main algorithm}
\label{code:intuition}
\end{figure}

@@

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. As explained in Proposition~\ref{prop:optimal-mln}, a simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $MLN_{ops}$. Then, assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = MLN_{ops} \cdot I$.

@@

One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$. In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities. By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient). Intuitively, $\iota$ captures this growth in non-hoistable operation.

@@

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$. In particular, we need to calculate the number terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators employed. For example, in the case of the bi-dimensional Poisson equation, after a suitable factorization, we have $\rho = 3$. There are several ways of determining $\rho$. The fastest would be to extract it from high-level analysis of the operators in the form; for convenience, in our implementation we instead project (by analysis of the expression tree) the output of expansion and factorization in the monomial.

@@

\section{Code Generation}
\label{sec:codegen}
The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake. In this section, we describe the structure of the code generation system and we also comment on a set of low-level optimizations.

@@

\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions. These are expressions that should logically be placed within $L_i$. They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point. For some complex monomials (e.g., see~\citeN{quadrature1} and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, outweigh that in the multilinear loop nest. In these cases, our definition of optimality becomes weaker, since the underlying assumption is that the bulk of the computation is carried out in innermost loops.

Despite the fact that we are not characterizing optimality for this wider class of loops, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis}, particularly for removing sharing, to minimize the operational cost. This is straightforward in our code generation system by composing rewrite operators. Our strategy is as follows.

COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions. This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to trivial code without performing any sort of transformation. Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$. We then synthesize an optimal loop nest as described in the previous sections. This may in turn expose a set of $L_i$-dependent statements evaluating some temporaries. For each of this statements, we try to remove sharing by greedily applying factorization and code motion. In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named "factorize", "expand", etc.).

@@
